---
title: "DV Assignment 2, Part 1: [Your Title Here]"
author: "Group X: Firstname Lastname, Firstname Lastname, Firstname Lastname"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

<!--
HOW TO KNIT THIS FILE

1. Open your group project in RStudio first (the `Project:` button in the
   top right should show your repo name). Keep this file in `report/`.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd in `report/`.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below sources scripts/00-packages.R and
   scripts/01-get-data.R, so packages and data are always loaded. The first
   knit downloads the data, so you need an internet connection.

If knitting fails:
- "could not find function": the package isn't loaded - add it to
  scripts/00-packages.R.
- "object not found": you created it in the Console, not in this file (or
  in a script that this file doesn't source).
- "cannot open file": check that the project is open, and use
  here("folder", "file") for paths instead of typing them out.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))
```

## Research question

*We have decided to look at how spreiding (within school variance of disatvantegness) could affect test scores at teacher advice across different levels of schoolweeging (school disadvantegeness). We though this would me interesting to see if more heterogenous groups with regards to advantaged/disadvantaged backround could me related to better performance. There is some evidence to suggest that students can benefit from more heterogenous backrounds so we though it interesting to try to see if we could see similarphenomenon in this data. We made several exploratory plots included in our test scripts folders but decided to keep only a couple of those here for simplicity. At this point the question is analytical, so we just look at the visualizations to see if there are any discernable patterns which we could later model. The audiance could be anyone interested in this data, but given the severe limitations of what we are measuring it would not be wise to draw big conclusion and present it to the wider public as a certain level nuance is required when interpreting the data.*

```{r variables_data, include = FALSE}
#Set up of the variables

# code. Average locations to institution level using pupil counts as weights.
weging_by_school <- schoolweging |>
  mutate(INSTELLINGSCODE = str_extract(OVT, "^[^|]+")) |>
  filter(
    !is.na(INSTELLINGSCODE),
    !is.na(schoolweging),
    !is.na(spreiding),
    !is.na(aantal_leerlingen),
    aantal_leerlingen > 0
  ) |>
  group_by(INSTELLINGSCODE) |>
  summarise(
    schoolweging = weighted.mean(schoolweging, aantal_leerlingen),
    spreiding = weighted.mean(spreiding, aantal_leerlingen),
    .groups = "drop"
  )

# STANDARDAZING TEST SCORES
# Provider is used only to standardize otherwise incomparable score scales.
scores_by_provider <- eindscores |>
  filter(SOORT_PO == "Bo") |>
  pivot_longer(
    cols = matches("^(IEP|ROUTE8|DIA|AMN|DOE|LIB)_(AANTAL|GEM)$"),
    names_to = c("provider", ".value"),
    names_pattern = "^(.+)_(AANTAL|GEM)$"
  ) |>
  mutate(
    AANTAL = as.numeric(
      if_else(
        str_detect(AANTAL, "^[0-9]+$"),
        AANTAL,
        NA_character_
      )
    )
  ) |>
  filter(AANTAL > 0, !is.na(GEM)) |>
  group_by(INSTELLINGSCODE, provider) |>
  summarise(
    mean_score = weighted.mean(GEM, AANTAL),
    tested_pupils = sum(AANTAL),
    .groups = "drop"
  ) |>
  group_by(provider) |>
  mutate(
    score_z = (mean_score - mean(mean_score)) / sd(mean_score)
  ) |>
  ungroup() |>
  filter(is.finite(score_z))

# Combine any multiple provider results into one school-level value, weighted
# by tested pupils. Provider names are not used in the visualizations.
scores_by_school <- scores_by_provider |>
  group_by(INSTELLINGSCODE) |>
  summarise(
    standardized_score = weighted.mean(score_z, tested_pupils),
    .groups = "drop"
  )
plot_data <- scores_by_school |>
  inner_join(weging_by_school, by = "INSTELLINGSCODE") |>
  filter(
    is.finite(standardized_score),
    is.finite(schoolweging),
    is.finite(spreiding)
  )

tertile_cutpoints <- quantile(
  plot_data$schoolweging,
  probs = c(1 / 3, 2 / 3),
  na.rm = TRUE
)
plot_data <- plot_data |>
  mutate(
    disadvantage_level = case_when(
      schoolweging <= tertile_cutpoints[[1]] ~ "Lower disadvantage",
      schoolweging <= tertile_cutpoints[[2]] ~ "Middle disadvantage",
      TRUE ~ "Higher disadvantage"
    ),
    disadvantage_level = factor(
      disadvantage_level,
      levels = c(
        "Lower disadvantage",
        "Middle disadvantage",
        "Higher disadvantage"
      )
    )
  )
```


##Exploratory process
```{r plot1, echo=FALSE}
plot_data3 <- scores_by_provider |>
  inner_join(weging_by_school, by = "INSTELLINGSCODE") |>
  filter(
    is.finite(mean_score),
    is.finite(schoolweging),
    is.finite(spreiding)
  )
plot_spread_and_weight <- ggplot(
  plot_data,
  aes(x = schoolweging, y = spreiding)
) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(
    title = "Schoolweging and spreiding",
    subtitle = "Each point represents one matched regular primary school",
    x = "Schoolweging (overall disadvantage)",
    y = "Spreiding (variation in disadvantage)"
  )
print(plot_spread_and_weight)
```
*The above graph just visualizes the variance of disatvanagness within school (y-axis) and different lvls of overall disadvantage of school (x-axis).Each point represent an individual institution. After seeing this we though it would be interesting to then see if you the schools in roughly the same schoolweging would have different performance scoresdepending on their spreiding.*

```{r plot2, echo=FALSE}

plot_score_by_spread <- ggplot(
  plot_data,
  aes(x = spreiding, y = standardized_score)
) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
  facet_wrap(vars(disadvantage_level)) +
  labs(
    title = "Relative scores by spreiding across disadvantage levels",
    subtitle = paste(
      "Schoolweging groups are data-derived tertiles;",
      "each point is one school"
    ),
    x = "Spreiding (variation in disadvantage)",
    y = "Standardized test score (provider-specific z-score)"
  )
print(plot_score_by_spread)
```
*The above graph uses standardized test scores(as schools use different providers) to see if student performance is higher with high or low variance of disadvantage.This is done for three lvls of overall disadvantagness: Low,Mid and High. There doesnt seem to be any apparent trend in the relationship we expected. However, we can also see that overall students with higher disadvantagness overall also have overall worse performance.*


```{r setup2, include = FALSE}
# ── 1. Schoolweging ───────────────────────────────────────────────────────────
# Extract INSTELLINGSCODE from OVT (format "00AP|C1") and aggregate to school
# level, weighting by number of pupils.
weging <- schoolweging |>
  mutate(INSTELLINGSCODE = str_extract(OVT, "^[^|]+")) |>
  filter(!is.na(schoolweging), !is.na(spreiding), aantal_leerlingen > 0) |>
  group_by(INSTELLINGSCODE) |>
  summarise(
    schoolweging = weighted.mean(schoolweging, aantal_leerlingen),
    spreiding    = weighted.mean(spreiding,    aantal_leerlingen),
    .groups = "drop"
  )

# ── 2. School advice → pct_havo_plus ─────────────────────────────────────────
# Replace "<5" with 2.5, convert to numeric, then compute the share of
# students advised for HAVO or higher.
parse_count <- function(x) {
  as.numeric(if_else(x == "<5", "2.5", x))
}

adviezen <- schooladviezen |>
  filter(SOORT_PO == "Bo") |>
  mutate(across(VSO:ADVIES_NIET_MOGELIJK, parse_count)) |>
  mutate(
    total     = rowSums(across(VSO:ADVIES_NIET_MOGELIJK), na.rm = TRUE),
    havo_plus = HAVO + HAVO_VWO + VWO
  ) |>
  filter(total > 0) |>
  mutate(pct_havo_plus = havo_plus / total) |>
  select(INSTELLINGSCODE, pct_havo_plus)

# ── 3. Join ───────────────────────────────────────────────────────────────────
plot_data2 <- adviezen |>
  inner_join(weging, by = "INSTELLINGSCODE") |>
  filter(
    is.finite(pct_havo_plus),
    is.finite(schoolweging),
    is.finite(spreiding)
  ) |>
  mutate(
    spreiding_q = cut(
      spreiding,
      breaks = quantile(spreiding, probs = 0:4 / 4, na.rm = TRUE),
      labels = c("Q1 (low spreiding)", "Q2", "Q3", "Q4 (high spreiding)"),
      include.lowest = TRUE
    )
  )

```



```{r plot3,echo=FALSE}
p3 <- ggplot(plot_data2, aes(x = schoolweging, y = pct_havo_plus)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  facet_wrap(vars(spreiding_q)) +
  labs(
    title = "Option 3: Faceted scatter by spreiding quartile",
    x = "Schoolweging", y = "Share advised HAVO or higher"
  )
print(p3)
```
*Additionaly, the same was performed but instead of standardized scores we use share of advised HAVO or higher. The plots are also this time split according to within school disadvantage variance rather than overall variance as was the case above.These visualization seem to follow a same pattern that there seems to be no effect of spreiding, but definitely an effect of schoolweging*
---

**Before submitting:** run the lintr check on this file, knit it, and push your work to GitHub.
