---
title: "DV Assignment 2: Effects of schools' socio-economic heterogeneity"
author: "Group 3: Filip Hlavinka, Henning Lang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
# Clean workspace:
rm(list = ls())

# Packages:
packages <- c("marginaleffects", "patchwork", "flextable")
for (i in packages) {
  if (!require(i, character.only = TRUE)) install.packages(i)
  library(i, character.only = TRUE)
  print(paste("Is package", i, "loaded?", i %in% loadedNamespaces()))
}
source(here::here("scripts", "00-packages.R"))

# Data
source(here::here("scripts", "01-get-data.R"))
```

## **Research question and target audience**

*RQ: How is within-school heterogeneity in socio-economic disadvantage associated with the proportion of students receiving final HAVO-or-higher advice, accounting for the school population’s average socio-economic disadvantage?* <br>

*Audience: Educational researchers interested in how socio-economic heterogeneity within schools relates to student performance, with attention to the limitations of causal interpretation and school-level data.*

## **Exploratory process**

### **Initial Data Preparation**

We combined the schoolweging (OVT column) and schooladviezen (INSTELLINGSCODE column) datasets using the school ID. The codebook did not explain how to match schools with several locations, so we excluded them. We replaced "<5" student counts with 2.5, as suggested in the codebook, and then created the HAVO+ variable, the share of final recommendations that are HAVO, HAVO/VWO or VWO (see explanation below).

```{r variable-prep, echo=FALSE, message=FALSE, warning=FALSE}
# Save school form variables
advice_cols <- c(
  "VSO", "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT",
  "VMBO_GT", "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)
havo_cols <- c("HAVO", "HAVO_VWO", "VWO")

# Keep institutions that appear exactly once in a source (single location);
# counted before dropping NAs, so empty locations still count as locations
keep_single <- function(df) {
  df |>
    add_count(INSTELLINGSCODE, name = "n_locations") |>
    filter(n_locations == 1L) |>
    select(-n_locations)
}

# "00AP|C1" -> "00AP"
school_context <- schoolweging |>
  transmute(
    INSTELLINGSCODE = str_remove(OVT, "\\|.*$"),
    schoolweging,
    spreiding,
    n_students_school = aantal_leerlingen
  ) |>
  keep_single()

# "<5" -> 2.5; share of advices that are HAVO or higher
school_advice <- schooladviezen |>
  select(INSTELLINGSCODE, all_of(advice_cols)) |>
  mutate(
    across(all_of(advice_cols), ~ as.numeric(str_replace(.x, "^<5$", "2.5"))),
    n_students_advice = rowSums(pick(all_of(advice_cols))),
    HAVO_higher = 100 * rowSums(pick(all_of(havo_cols))) / n_students_advice
  ) |>
  select(INSTELLINGSCODE, HAVO_higher, n_students_advice) |>
  keep_single()

# One row per school
school_data <- school_context |>
  inner_join(school_advice, by = "INSTELLINGSCODE", 
             relationship = "one-to-one") |>
  filter(
    !is.na(schoolweging), !is.na(spreiding),
    !is.na(HAVO_higher), n_students_advice > 0
  )

head(school_data)
```

In the joined dataset the following variables are particularly important for our visualization:

| Variable name in data | Original dataset | Variable name in plots | Explanation | Predictor/Outcome |
|---|---|---|-------------|
| `spreiding` | `schoolweging` | Socio-economic Heterogeneity | Differences in socio-economic disadvantage among pupils within a school; higher values indicate greater heterogeneity. |  **Predictor** |
| `schoolweging` | `schoolweging` | Avg. socio-economic disadvantage | Measure of the mean school population’s socio-economic disadvantage; higher values indicate greater disadvantage. | **Predictor** |
| `HAVO_higher` | Derived from `schooladviezen` | HAVO+ advice | Percentage of recorded final recommendations that are HAVO, HAVO/VWO, or VWO | **Outcome** |

### **First plot on Havo+ advice**

```{r first-plot, echo=FALSE, fig.width = 12, fig.height = 6, fig.align = "center"}
# ------------------------------------------------------------------------------
# First Exploratory plot

# Prepare the data for the first exploratory plot:
advice_share <- schooladviezen |>
  semi_join(school_data, by = "INSTELLINGSCODE") |>
  select(all_of(advice_cols)) |>
  mutate(across(everything(), ~ as.numeric(str_replace(.x, "^<5$", "2.5")))) |>
  summarise(across(everything(), sum)) |>
  pivot_longer(everything(), names_to = "advice", values_to = "n") |>
  mutate(
    share = 100 * n / sum(n),
    havo_plus = advice %in% havo_cols,
    advice = factor(
      str_replace_all(advice, "_", "-"),
      levels = str_replace_all(advice_cols, "_", "-")
    )
  )

# Create the first exploratory plot:
ggplot(advice_share, aes(x = advice, y = share, fill = havo_plus)) +
  geom_col() +
  geom_text(
    aes(label = sprintf("%.1f%%", share)),
    vjust = -0.4, fontface = "bold"
  ) +
  # cut between VMBO-GT-HAVO (8th bar) and HAVO (9th bar)
  geom_vline(xintercept = 8.5, linetype = "dashed", colour = "grey30") +
  # arrow and label above the three HAVO-or-higher bars
  annotate(
    "segment",
    x = 8.6, xend = 11.4, y = 24, yend = 24,
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    colour = "#2c7fb8"
  ) +
  annotate(
    "text",
    x = 10, y = 22.8, label = "Final advice for HAVO or higher\n(our outcome variable)",
    colour = "#2c7fb8", fontface = "bold"
  ) +
  scale_fill_manual(values = c("TRUE" = "#74add1", "FALSE" = "Medium Spring Green")) +
  scale_y_continuous(
    breaks = seq(0, 20, 5), limits = c(0, 25),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    title = "Final school advice across all Dutch primary schools",
    subtitle = "Share of pupils per secondary-school track, 2024-25",
    x = "Final school advice",
    y = "Share of pupils (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

<u>Description</u>

We placed the cut between VMBO-GT/HAVO and HAVO to obtain two roughly similar-sized groups of recommendations. This also follows the codebook’s track ordering: HAVO, HAVO/VWO, and VWO represent the higher academic recommendations, and we calculate their combined share within each school.

### **Second plot on bivariate relationships**

```{r second-plot, echo=FALSE}
# ------------------------------------------------------------------------------
# Second exploratory plot

# Scatterplot: points sized by number of students per school
scatter <- function(x, y, x_lab, y_lab) {
  ggplot(school_data, aes({{ x }}, {{ y }}, size = n_students_school)) +
    geom_point(alpha = 0.3, colour = "#2c7fb8") +
    scale_size_continuous(name = "Students per school", range = c(0.3, 3.5)) +
    guides(size = guide_legend(override.aes = list(alpha = 0.7))) +
    labs(x = x_lab, y = y_lab)
}

# Titles and labs
het <- "Socio-economic heterogeneity"
dis <- "Average socio-economic disadvantage"
adv <- "HAVO or higher advice (%)"

# Create second plot with patchwork
(scatter(spreiding, HAVO_higher, het, adv) +
    scatter(schoolweging, HAVO_higher, dis, adv) +
    scatter(schoolweging, spreiding, dis, het)) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Exploring relationships between socio-economic school context and HAVO advice",
    subtitle = "Two-variable relationships; each dot is one school"
  ) &
  theme_minimal() &
  theme(legend.position = "bottom")
```

<u>Description</u>

The relationship between socio-economic heterogeneity and HAVO+ advice is not overt, whereas avg. socio-economic disadvantage appears to have a negative linear relationship with HAVO+ advice. Avg. Disadvantage and heterogeneity appear to follow an inverted-U-shaped.

```{r second-prep, echo=FALSE}
# ------------------------------------------------------------------------------
# Subsequent data preparation

# Investigatiion of schoolweging based on the mean of spreiding:
school_data |>
  mutate(decile = ntile(schoolweging, 10)) |>
  summarise(
    .by = decile,
    min_weging = mean(schoolweging),
    mean_spreiding = mean(spreiding), n = n()
  ) |>
  arrange(decile)

# Keep all schools for the plot that explains the trimming
school_data_full <- school_data

# Trimming of schoolweging (cutting the lowest and highest 10%),
# and categorize the rest of schoolweging into 10 bins with same sample size.
# Centered the continuous varables for the subsequent interaction models.
limits <- quantile(school_data$schoolweging, c(1 / 10, 9 / 10))
school_data <- school_data |>
  filter(between(schoolweging, limits[[1]], limits[[2]])) |>
  mutate(
    weging_trimmed = schoolweging,
    weging_bin = factor(ntile(weging_trimmed, 10), ordered = TRUE),
    weging_trim_c = weging_trimmed - mean(weging_trimmed),
    spreiding_c = spreiding - mean(spreiding)
  )

head(school_data)
```

<u>Description</u>

...

```{r third-plot, echo=FALSE}
# ------------------------------------------------------------------------------
# Third exploratory plot

bin_colours <- c(
  "1"  = "#071440",  # deep navy
  "2"  = "#10286E",  # dark royal blue
  "3"  = "#1A459F",  # royal blue
  "4"  = "#2563C4",  # strong blue
  "5"  = "#3F86DB",  # sky blue 
  "6"  = "#4A9C8A",  # muted teal
  "7"  = "#5BA777",  # muted green
  "8"  = "#78B067",  # soft green
  "9"  = "#9BB95E",  # olive green
  "10" = "#BCC45C"   # muted yellow-green
)

bin_info <- school_data |>
  summarise(
    .by = weging_bin,
    lo = min(schoolweging), hi = max(schoolweging), n_schools = n()
  ) |>
  mutate(
    x = as.integer(weging_bin),
    txt = if_else(x <= 6, "white", "black"),
    range = paste(round(lo, 1), round(hi, 1), sep = "-")
  )

bin_line_colour      <- "grey45"
trim_line_colour     <- "red"
outside_point_colour <- "grey75"

# Preserve the full dataset BEFORE trimming
bin_labels <- setNames(
  paste0(
    "Bin ", bin_info$weging_bin,
    " (n=", bin_info$n_schools, ")"
  ),
  as.character(bin_info$weging_bin)
)

# Place boundaries halfway between adjacent ntile() bins
bin_boundaries <- school_data |>
  group_by(weging_bin) |>
  summarise(
    bin_min = min(schoolweging),
    bin_max = max(schoolweging),
    .groups = "drop"
  ) |>
  arrange(weging_bin) |>
  mutate(cutoff = (bin_max + lead(bin_min)) / 2) |>
  filter(!is.na(cutoff))

plot_spread_and_weight2 <- ggplot(
  school_data_full,
  aes(x = schoolweging, y = spreiding)
) +
  # Full dataset in grey
  geom_point(
    colour = outside_point_colour,
    alpha = 0.5,
    size = 0.5
  ) +
  # Only schools within the trimming limits get bin colours
  geom_point(
    data = school_data,
    aes(colour = weging_bin),
    alpha = 0.5,
    size = 0.5
  ) +
  # Boundaries between ntile() bins
  geom_vline(
    data = bin_boundaries,
    aes(xintercept = cutoff),
    colour = bin_line_colour,
    linetype = "dotted",
    linewidth = 0.4
  ) +
  # 10th- and 90th-percentile limits
  geom_vline(
    xintercept = limits,
    colour = trim_line_colour,
    linetype = "dashed",
    linewidth = 0.6
  ) +
  scale_colour_manual(
    values = bin_colours,
    labels= bin_labels,
    name = "Schoolweging bin",
    drop = FALSE
  ) +
  guides(
    colour = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      ncol = 5,              # Fewer entries per row prevent crowding
      byrow = TRUE,
      override.aes = list(
        size = 2,            # Make legend dots easier to see
        alpha = 1
      )
    )
  ) +
  labs(
    title = "Average socio-economic disadvantage (Schoolweging) vs. 
      Within-school socioeconomic heterogeneity (Spreiding)",
    subtitle = "",
    x = "Avg. socio-economic disadvantage",
    y = "Socio-economic heterogeneity"
  ) +
  theme_minimal() +
  # Put custom settings AFTER theme_minimal()
  theme(
    plot.title = element_text(
      face = "bold",
      size = 12,
      hjust = 0.5),
    plot.title.position = "plot",
    plot.margin = margin(0, 5.5, 5.5, 5.5),
    axis.title = element_text(size = 10),
    strip.text = element_text(face = "bold"),
    
    legend.position = "top",
    legend.title = element_text(size = 8, face = "italic"),
    legend.text = element_text(size = 7),
    legend.key.width = grid::unit(0.2, "cm"),
    legend.key.height = grid::unit(0.2, "cm")
  )

plot_spread_and_weight2
```

<u>Description</u>
We look at simple scatter plot of Avg. socio-economic disadvantage and socioeconomic heterogenity within a school. Cutoff points of 10% on each side of the graph were made (red lines) and the schools were excluded. This was done because we wanted to compare schools in roughly the same avg. socio-economic disadvantage and unfortunately at both ends schools with similar avg. socio-economic disadvantage and differing heterogenity levels are not so abundant. Additionally this was done to focus more on middle range of avg. disadvantaged schools 25-35. Furthermore, this range was split into 10 bins with the same size of schools ~445. We then use these bins to see how heterogenity is associated with havo+ advice across different bins.

## **Statistical MOdelling**

```{r regressions, echo=FALSE,}
# ------------------------------------------------------------------------------
# Regressions

# Model 1
model_1 <- lm(
  HAVO_higher ~ spreiding_c * weging_bin, data = school_data)
summary(model_1)

# Model 2
model_2 <- lm(HAVO_higher ~ spreiding_c * weging_trim_c, data = school_data)
summary(model_2)

# Check regression assumptions for models
old_par <- par(mfrow = c(2, 2))
plot(model_1, which = c(1, 2, 3, 5), ask = FALSE)
par(old_par)
# old_par <- par(mfrow = c(2, 2))
# plot(model_2, which = c(1, 2, 3, 5), ask = FALSE)
# par(old_par)
# --> Similar pattern for both models, so model 1 is chosen.
# --> Non-normality at the tails can be seen. 
```

## **Final visualization**

*Your final, explanatory visualization goes below. Include a short write-up: what does it communicate, why is this the best way to communicate it, and how robust is the finding?*

```{r final-plot, fig.width = 14, fig.height = 8.5, dpi = 200, out.width = "100%", fig.align = "center", fig.alt = "Slopes of socio-economic heterogeneity on HAVO+ advice per disadvantage bin", echo = FALSE, warning = FALSE, message = FALSE}
# ------------------------------------------------------------------------------
# Final visualization - shared settings

# Labels for the axis
het <- "Socio-economic heterogeneity"
adv <- "HAVO+ advice"
dis <- "Avg. socio-economic disadvantage"

# Shared legend style, so that all legends have the same text and key size
legend_theme <- theme(
  legend.position = "bottom",
  legend.title = element_text(size = 7.5),
  legend.text = element_text(size = 7.5),
  legend.key.height = unit(0.5, "cm"),
  legend.box.spacing = unit(4, "pt")
)

# ------------------------------------------------------------------------------
# Final visualization plot - bottom left

# Bin-specific slopes with conventional 95% CIs
slope_data <- marginaleffects::avg_slopes(
  model_1, variables = "spreiding_c", by = "weging_bin"
)

# Fitted lines with 95% bands over each bin's observed range
spreiding_mean <- mean(school_data$spreiding)
pred_data <- school_data |>
  reframe(
    .by = weging_bin,
    spreiding = seq(min(spreiding), max(spreiding), length.out = 50)
  ) |>
  mutate(spreiding_c = spreiding - spreiding_mean) |>
  marginaleffects::predictions(model = model_1, newdata = _)

# Mean advice per bin: a flat line at this level means "no relationship"
bin_means <- school_data |>
  summarise(.by = weging_bin, mean_advice = mean(HAVO_higher))

# Legend text for the dashed lines (used in both bottom plots)
zero_label <- "No relationship (zero slope)"

# Create the bottom left plot:
p_left <- ggplot(school_data, aes(x = spreiding, y = HAVO_higher)) +
  geom_point(
    aes(size = n_students_school),
    alpha = 0.15, colour = "grey35"
  ) +
  geom_hline(
    data = bin_means,
    aes(yintercept = mean_advice, linetype = zero_label),
    colour = "grey30", linewidth = 0.5
  ) +
  geom_ribbon(
    data = pred_data,
    aes(
      y = estimate, ymin = conf.low, ymax = conf.high,
      fill = weging_bin
    ),
    alpha = 0.25
  ) +
  geom_line(
    data = pred_data,
    aes(y = estimate, colour = weging_bin),
    linewidth = 0.9
  ) +
  facet_wrap(
    ~ weging_bin, ncol = 5,
    labeller = labeller(weging_bin = ~ paste("Bin", .x))
  ) +
  scale_colour_manual(values = bin_colours, guide = "none") +
  scale_fill_manual(values = bin_colours, guide = "none") +
  scale_size_continuous(
    name = paste(
      "Students per school:",
      "Models unweighted; point size for illustration"
    ),
    range = c(0.3, 2.5)
  ) +
  scale_linetype_manual(name = NULL, values = "dashed") +
  guides(
    size = guide_legend(
      override.aes = list(alpha = 0.7),
      theme = theme(legend.title.position = "left")
    ),
    linetype = guide_legend(
      theme = theme(legend.key.width = unit(1.2, "cm"))
    )
  ) +
  scale_y_continuous(
    breaks = seq(0, 100, 25),
    labels = scales::label_percent(scale = 1)
  ) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    title = "What the data look like",
    subtitle = "Each dot is one school; line = linear fit with 95% CI",
    x = het, y = adv
  ) +
  theme_minimal() +
  legend_theme +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.margin = margin(0, 5.5, 5.5, 5.5),
    axis.title = element_text(size = 10),
    strip.text = element_text(face = "bold")
  )

# ------------------------------------------------------------------------------
# Final visualization plot - bottom right

# Data for the dashed zero line, so that it gets its own legend entry
zero_line <- data.frame(
  yintercept = 0,
  explanation = "If CI includes zero = slope not significant"
)

# Create the bottom right plot:
p_right <- ggplot(
  slope_data,
  aes(x = weging_bin, y = estimate, colour = weging_bin)
) +
  geom_hline(
    data = zero_line,
    aes(yintercept = yintercept, linetype = explanation),
    colour = "grey50", inherit.aes = FALSE
  ) +
  geom_pointrange(
    aes(ymin = conf.low, ymax = conf.high),
    linewidth = 0.8, size = 0.6
  ) +
  geom_text(
    aes(label = sprintf("%+.1f pp", estimate)),
    colour = "black", fontface = "bold",
    hjust = 0, nudge_x = 0.15, size = 2.7
  ) +
  scale_x_discrete(expand = expansion(add = c(0.6, 1.2))) +
  scale_colour_manual(values = bin_colours, guide = "none") +
  scale_linetype_manual(name = NULL, values = "dashed") +
  guides(
    linetype = guide_legend(
      theme = theme(legend.key.width = unit(1.2, "cm"))
    )
  ) +
  scale_y_continuous(
    breaks = seq(-2, 6, 2),
    labels = scales::label_number(style_positive = "plus", suffix = " pp")
  ) +
  coord_cartesian(ylim = c(-2, 6)) +
  labs(
    title = "Slopes of socio-economic heterogeneity on advice",
    subtitle = paste(
      "Slope estimated separately within each disadvantage bin;",
      "95% CI"
    ),
    x = paste(dis, "(Bin)"),
    y = paste0("Change in ", adv, " per unit of ", tolower(het))
  ) +
  theme_minimal() +
  legend_theme +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.margin = margin(0, 5.5, 5.5, 5.5),
    axis.title = element_text(size = 10)
  )

# ------------------------------------------------------------------------------
# Final visualization - plot title and subtitle at the top:

# Define title and subtitle:
final_title <- paste(
  "More mixed school populations, higher advice:",
  "the association is clearest where disadvantage is lower"
)
final_subtitle <- paste(
  "How is within-school heterogeneity in socio-economic disadvantage",
  "associated with the proportion of students receiving final",
  "HAVO-or-higher advice, accounting for the school population’s",
  "average socio-economic disadvantage?"
)

# ------------------------------------------------------------------------------
# Final visualization legend bins

# Height of the legend panel in cm (fixed; used in plot_layout() below)
top_cm <- 2.7
# Distance in cm between the legend box and the plots below
gap_cm <- 0.1
# Approximate cm per y-unit of the panel (y runs from -1 to 1)
cm_per_y <- (top_cm - 0.15) / 2

# Strip geometry (x: 0-100 across the panel; y: -1 to 1)
strip_x0 <- 44.5
tile_w <- 5.2
tile_h <- 0.76
tile_yc <- 0.13
y_head <- 0.66
y_top <- 0.78
y_bottom <- -0.78
# "Bin X" size equals the facet strip titles in the left plot (8.8 pt)
bin_font <- 8.8 / ggplot2::.pt

# Bin summary: range of average disadvantage and number of schools per bin
bin_info <- school_data |>
  summarise(
    .by = weging_bin,
    lo = min(schoolweging), hi = max(schoolweging), n_schools = n()
  ) |>
  mutate(
    x = as.integer(weging_bin),
    tile_x = strip_x0 + (x - 0.5) * tile_w,
    txt = if_else(x <= 6, "white", "black"),
    range = paste(round(lo, 1), round(hi, 1), sep = "-")
  )

# Two note rows below the strip: bin ranges, schools per bin, trimming
strip_note <- paste0(
  "Numbers within bins show the range of avg. socio-economic disadvantage. ",
  paste(unique(range(bin_info$n_schools)), collapse = "-"),
  " schools per bin.\n",
  "Schools in the lowest and highest disadvantage deciles excluded ",
  "due to lower within-school socio-economic heterogeneity."
)

# Explanation box: bold term (left column) + short sentence (right column)
box_font <- 3
box_items <- data.frame(
  term = c(
    "HAVO+ advice:",
    "Avg. socio-economic\ndisadvantage:",
    "Socio-economic\nheterogeneity:"
  ),
  text = c(
    "Share of final recommendations that are HAVO, HAVO/VWO or VWO.",
    paste(
      "Average socio-economic disadvantage of the school population;",
      "higher = more disadvantaged. Defines the bins."
    ),
    paste(
      "Differences in socio-economic disadvantage among pupils",
      "within a school; higher = more heterogeneous."
    )
  )
)
box_items$text <- stringr::str_wrap(box_items$text, width = 60)

# Spread the items evenly over the same height as the right-hand strip
line_h <- box_font * ggplot2::.pt * 1.2 * 0.95 / 72 * 2.54 / cm_per_y
box_items$n_lines <- pmax(
  stringr::str_count(box_items$text, "\n"),
  stringr::str_count(box_items$term, "\n")
) + 1
text_h <- sum(box_items$n_lines) * line_h
item_gap <- max((y_top - y_bottom - text_h) / (nrow(box_items) - 1), 0.04)
box_items$y <- y_top -
  c(0, cumsum(head(box_items$n_lines * line_h + item_gap, -1)))

# Create the legend panel: one box around explanation and bin strip
p_legend <- ggplot(bin_info) +
  annotate(
    "rect",
    xmin = 1.5, xmax = 98.5, ymin = -0.94, ymax = 0.94,
    fill = "grey97", colour = "grey30", linewidth = 0.3
  ) +
  # explanation text
  geom_text(
    data = box_items,
    aes(x = 3.5, y = y, label = term),
    hjust = 0, vjust = 1, fontface = "bold", size = box_font,
    lineheight = 0.95, colour = "grey15"
  ) +
  geom_text(
    data = box_items,
    aes(x = 14.5, y = y, label = text),
    hjust = 0, vjust = 1, size = box_font, lineheight = 0.95,
    colour = "grey15"
  ) +
  # compressed bin strip
  geom_tile(
    aes(x = tile_x, y = tile_yc, fill = weging_bin),
    width = tile_w * 0.94, height = tile_h
  ) +
  geom_text(
    aes(
      x = tile_x, y = tile_yc + 0.15,
      label = paste("Bin", x), colour = txt
    ),
    fontface = "bold", size = bin_font
  ) +
  geom_text(
    aes(x = tile_x, y = tile_yc - 0.15, label = range, colour = txt),
    size = 2.6
  ) +
  annotate(
    "text", x = strip_x0 + 5 * tile_w, y = y_head, fontface = "bold",
    size = 3.3, colour = "grey15",
    label = "10 equal-sized bins of avg. socio-economic disadvantage"
  ) +
  annotate(
    "text", x = strip_x0, y = y_head, hjust = 0,
    size = 3.1, colour = "grey15", label = "\u2190"
  ) +
  annotate(
    "text", x = strip_x0 + 2.2, y = y_head, hjust = 0,
    fontface = "italic", size = box_font, colour = "grey15",
    label = "Less disadvantaged"
  ) +
  annotate(
    "text", x = strip_x0 + 10 * tile_w, y = y_head, hjust = 1,
    size = 3.1, colour = "grey15", label = "\u2192"
  ) +
  annotate(
    "text", x = strip_x0 + 10 * tile_w - 2.2, y = y_head, hjust = 1,
    fontface = "italic", size = box_font, colour = "grey15",
    label = "More disadvantaged"
  ) +
  # two note rows: bin ranges, schools per bin, trimming
  annotate(
    "text", x = strip_x0 + 5 * tile_w, y = -0.58, size = 2.7,
    lineheight = 0.95, colour = "grey30", label = strip_note
  ) +
  scale_fill_manual(values = bin_colours) +
  scale_colour_identity() +
  coord_cartesian(xlim = c(0, 100), ylim = c(-1, 1), expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(2, 0, 2, 0)
  )

# ------------------------------------------------------------------------------
# Final plot

# Bottom row: scatterplot (left), slope plot (right)
bottom_row <- (p_left | p_right) + plot_layout(widths = c(2.7, 2.3))

# Create final plot: the legend box is centred, a spacer sets the distance
# between the box and the plots
final_plot <- wrap_elements(full = p_legend) / plot_spacer() / bottom_row +
  plot_layout(heights = unit(c(top_cm, gap_cm, 1), c("cm", "cm", "null"))) +
  plot_annotation(
    title = final_title,
    subtitle = final_subtitle,
    theme = theme(
      plot.title = element_text(face = "bold", size = 18),
      plot.subtitle = element_text(
        colour = "grey30", size = 9.2, margin = margin(b = 13)
      )
    )
  )
final_plot

ggsave(
  "final_plot.pdf",
  plot   = final_plot,
  width  = 14,
  height = 8.5,
  units  = "in",
  device = cairo_pdf
)
```

## **Conclusion**
<u></u>
Conclusion: 
Within the middle range of average socio-economic disadvantage that we examine, greater within-school heterogeneity is associated with a higher proportion of students receiving final HAVO-or-higher advice, mainly among relatively less disadvantaged schools (bins 1–5). In these bins, a one-unit increase in heterogeneity is associated with approximately 3.5–2.5 (bins 1-5 respectively) percentage points more HAVO+ advice. The association is non existent or generally weaker in more disadvantaged schools, although it remains positive in bin 10. These findings suggest that the relationship between socio-economic heterogeneity and educational advice varies with schools average disadvantage. Important to mention that this is definitely not evidence for a causal effect.
<br>
Limitations: While our analysis identifies a relationship, the estimated  effect sizes and the difference between  bins are not, on their own, sufficient  to support policy recommendations. Furthermore, school-level heterogeneity may also obscure classroom-level effect, where mechanisms such as peer interactions and teaching practices are likely to have a  more direct effect. These mechanisms are complex and may depend on how students are grouped and supported; a heterogeneous school population does not necessarily imply heterogeneous classrooms. Our findings therefore do not establish that increasing heterogeneity would improve outcomes. Nevertheless, our findings are consistent with existing body of knowledge socio-economic diversity can be beneficial to student performancce under certain conditions.


<u>Description of AI use</u>
Filip: I have used AI mainly for exploratory part of the assignment. So to check how the data looks, what might be point of interests etc. I have used copilot for those plots. Since it did not provide the best looking plot at first manual changes were then made to change to plot appropriatly of how we wanted it to look. Also used to look up arguments and if I wanted specific changes in a plot.
<br>
Henning:
---

**Before submitting:** run the lintr check on this file, and re-knit from a clean environment (`rm(list = ls())`, already done above) to make sure everything reproduces from scratch.
