---
title: "DV Assignment 2: Effects of schools' socio-economic heterogeneity"
author: "Group 3: Filip Hlavinka, Henning Lang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
# Clean workspace:
rm(list = ls())

# Packages:
packages <- c("marginaleffects", "patchwork", "kableExtra")
# --> marginaleffects for bin-specific regression slopes and CIs
# --> patchwork for combination of ggplots
# --> kableExtra for beautiful markdown tables
for (i in packages) {
  if (!require(i, character.only = TRUE)) install.packages(i)
  library(i, character.only = TRUE)
  print(paste("Is package", i, "loaded?", i %in% loadedNamespaces()))
}
source(here::here("scripts", "00-packages.R"))

# Data
source(here::here("scripts", "01-get-data.R"))
```

## **Research question and audience**

<u>Research question</u>

We examine whether schools with more socio-economically heterogeneous populations differ in their share of HAVO-or-higher recommendations, comparing schools with similar average levels of disadvantage. We would like to explore this question without a directional hypothesis, using public dutch school-level data for 2024/2025. Given the data, we examine associations rather than causal effects. Our research question is:

> **How is within-school heterogeneity in socio-economic disadvantage associated with the proportion of students receiving final HAVO-or-higher advice, accounting for the school population’s average socio-economic disadvantage?**

Note: Socio-economic disadvantage was measured ...

<u>Level of visualization</u>

Our chosen level of visualization is **inference**. 

<u>Target audience</u>

Our visualization is intended as a **research poster for educational researchers** interested in how socio-economic differences among pupils and schools relate to final secondary-school recommendations. The graphic (in poster form) should, for example, be usable for presenting results at a scientific conference. The interpretation of the visualization requires basic statistical knowledge, e. g. familiarity with linear regression, interaction terms, and confidence intervals, allowing readers to compare associations across schools with different levels of average disadvantage. Policymakers with a research background and an interest in school composition may also find it relevant, provided they recognize the limits of school-level data and non-causal findings.

## **Exploratory process**

### Initial Data Preparation

We combined the schoolweging (OVT column) and schooladviezen (INSTELLINGSCODE column) datasets using the school ID. The codebook did not explain how to match schools with several locations, so we excluded them. We replaced "<5" student counts with 2.5, as suggested in the codebook, and then created the HAVO+ variable, the share of final recommendations that are HAVO, HAVO/VWO or VWO (see explanation below).

After this initial preparation, the data looked like this:

```{r variable-prep, echo=FALSE, message=FALSE, warning=FALSE}
# Save school form variables
advice_cols <- c(
  "VSO", "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT",
  "VMBO_GT", "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)
havo_cols <- c("HAVO", "HAVO_VWO", "VWO")

# Keep institutions that appear exactly once in a source (single location);
# counted before dropping NAs, so empty locations still count as locations
keep_single <- function(df) {
  df |>
    add_count(INSTELLINGSCODE, name = "n_locations") |>
    filter(n_locations == 1L) |>
    select(-n_locations)
}

# "00AP|C1" -> "00AP"
school_context <- schoolweging |>
  transmute(
    INSTELLINGSCODE = str_remove(OVT, "\\|.*$"),
    schoolweging,
    spreiding,
    n_students_school = aantal_leerlingen
  ) |>
  keep_single()

# "<5" -> 2.5; share of advices that are HAVO or higher
school_advice <- schooladviezen |>
  select(INSTELLINGSCODE, all_of(advice_cols)) |>
  mutate(
    across(all_of(advice_cols), ~ as.numeric(str_replace(.x, "^<5$", "2.5"))),
    n_students_advice = rowSums(pick(all_of(advice_cols))),
    HAVO_higher = 100 * rowSums(pick(all_of(havo_cols))) / n_students_advice
  ) |>
  select(INSTELLINGSCODE, HAVO_higher, n_students_advice) |>
  keep_single()

# One row per school
school_data <- school_context |>
  inner_join(school_advice, by = "INSTELLINGSCODE", 
             relationship = "one-to-one") |>
  filter(
    !is.na(schoolweging), !is.na(spreiding),
    !is.na(HAVO_higher), n_students_advice > 0
  )

kable(head(school_data)) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```

In the joined dataset the following variables are particularly important for our visualization:

| Variable name in data | Original dataset | Variable name in plots | Explanation | Predictor/Outcome |
|---|---|---|-------------|
| `spreiding` | `schoolweging` | Socio-economic Heterogeneity | Differences in socio-economic disadvantage among pupils within a school; higher values indicate greater heterogeneity. |  **Predictor** |
| `schoolweging` | `schoolweging` | Avg. socio-economic disadvantage | Measure of the mean school population’s socio-economic disadvantage; higher values indicate greater disadvantage. | **Predictor** |
| `HAVO_higher` | `schooladviezen` | HAVO+ advice | Percentage of recorded final recommendations that are HAVO, HAVO/VWO, or VWO | **Outcome** |

### Plot 1 on Havo+ advice

With this plot we want to explain how we created the outome HAVO+ advice:

```{r first-plot, echo=FALSE, fig.align="center", fig.height=6, fig.width=12, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# First plot

# Prepare the data for the first exploratory plot:
advice_share <- schooladviezen |>
  semi_join(school_data, by = "INSTELLINGSCODE") |>
  select(all_of(advice_cols)) |>
  mutate(across(everything(), ~ as.numeric(str_replace(.x, "^<5$", "2.5")))) |>
  summarise(across(everything(), sum)) |>
  pivot_longer(everything(), names_to = "advice", values_to = "n") |>
  mutate(
    share = 100 * n / sum(n),
    havo_plus = advice %in% havo_cols,
    advice = factor(
      str_replace_all(advice, "_", "-"),
      levels = str_replace_all(advice_cols, "_", "-")
    )
  )

# Create the first exploratory plot:
ggplot(advice_share, aes(x = advice, y = share, fill = havo_plus)) +
  geom_col() +
  geom_text(
    aes(label = sprintf("%.1f%%", share)),
    vjust = -0.4, fontface = "bold"
  ) +
  geom_vline(xintercept = 8.5, linetype = "dashed", colour = "grey30") +
  annotate(
    "segment",
    x = 8.6, xend = 11.4, y = 24, yend = 24,
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    colour = "#2c7fb8"
  ) +
  annotate(
    "text",
    x = 10, y = 22.8, label = "Final advice for HAVO or higher\n(our outcome variable)",
    colour = "#2c7fb8", fontface = "bold"
  ) +
  scale_fill_manual(values = c("TRUE" = "#74add1", "FALSE" = "Medium Spring Green")) +
  scale_y_continuous(
    breaks = seq(0, 20, 5), limits = c(0, 25),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    title = "Final school advice across all Dutch primary schools",
    subtitle = "Share of pupils per secondary-school track, 2024-25",
    x = "Final school advice",
    y = "Share of pupils (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

<u>Description</u>

We placed the cut between VMBO-GT/HAVO and HAVO to obtain two roughly similar-sized groups of recommendations. This also follows the codebook’s track ordering: HAVO, HAVO/VWO, and VWO represent the higher academic recommendations, and we calculate their combined share within each school.

### Plot 2 on bivariate relationships

In this exploratory plot we look at the bivariate relations:

```{r second-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.align="center", fig.height=6, fig.width=12}
# ------------------------------------------------------------------------------
# Second plot

# Scatterplot: points sized by number of students per school
scatter <- function(x, y, x_lab, y_lab) {
  ggplot(school_data, aes({{ x }}, {{ y }}, size = n_students_school)) +
    geom_point(alpha = 0.3, colour = "#2c7fb8") +
    scale_size_continuous(name = "Students per school", range = c(0.3, 3.5)) +
    guides(size = guide_legend(override.aes = list(alpha = 0.7))) +
    labs(x = x_lab, y = y_lab)
}

# Titles and labs
het <- "Socio-economic heterogeneity"
dis <- "Avg. socio-economic disadvantage"
adv <- "HAVO+ advice (%)"

# Create second plot with patchwork
(scatter(spreiding, HAVO_higher, het, adv) +
    scatter(schoolweging, HAVO_higher, dis, adv) +
    scatter(schoolweging, spreiding, dis, het)) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Exploring relationships between socio-economic school context and HAVO advice",
    subtitle = "Two-variable relationships; each dot is one school",
    theme = theme_minimal()
  ) &
  theme_minimal() &
  theme(legend.position = "bottom")
```

<u>Description</u>

The relationship between socio-economic heterogeneity and HAVO+ advice is not overt, whereas avg. socio-economic disadvantage appears to have a negative linear relationship with HAVO+ advice. Avg. Disadvantage and heterogeneity appear to follow an inverted-U-shaped.

### Subsequent Data Preparation

Because of the quadratic relation between avg. disadvantage and heterogeneity we looked at the mean heterogeneity within the deciles of avg. disadvantage:

```{r avg_da_deciles, echo=FALSE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# Deciles of avg. disadvantage and mean heterogeneity 

# Investigatiion of schoolweging based on the mean of spreiding:
deciles <- school_data |>
  mutate(decile = ntile(schoolweging, 10)) |>
  summarise(
    .by = decile,
    mean_weging = mean(schoolweging),
    mean_spreiding = mean(spreiding),
    n = n()
  ) |> 
  arrange(decile)

kable(deciles, digits = 2) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

It can be seen that the mean heterogeneity for the lowest and highest avg. disadvantage deciles is substantially lower than for the other deciles. Therefore we excluded the lowest and highest deciles. Also, for the interaction models we centred the continuous avg. disadvantage variable and the heterogeneity variable. 

After this second data preparation the data looked like this (here only the most important variables): 

```{r second-prep, echo=FALSE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# Subsequent data preparation

# Keep all schools for the plot that explains the trimming
school_data_full <- school_data

# Trimming of schoolweging (cutting the lowest and highest 10%),
# and categorize the rest of schoolweging into 10 bins with same sample size.
# Centred the continuous variables for the subsequent interaction models.
limits <- quantile(school_data$schoolweging, c(1 / 10, 9 / 10))
school_data <- school_data |>
  filter(between(schoolweging, limits[[1]], limits[[2]])) |>
  mutate(
    weging_trimmed = schoolweging,
    weging_bin = factor(ntile(weging_trimmed, 10), ordered = TRUE),
    weging_trim_c = weging_trimmed - mean(weging_trimmed),
    spreiding_c = spreiding - mean(spreiding)
  )

kable(head(school_data |> 
             select(INSTELLINGSCODE, n_students_school, 
                    HAVO_higher, weging_trim_c,
                    weging_bin, spreiding_c))) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```

### Plot 3 on trimmed avg. disadvantage

In this plot we want to explain how we trimmed and grouped the avg. disadvantage:

```{r third-plot, echo=FALSE, fig.width=12, fig.height=6, fig.align="center"}
# ------------------------------------------------------------------------------
# Third plot

# Define colours for all subsequent plots
bin_colours <- c(
  "1"  = "#071440",  # deep navy
  "2"  = "#10286E",  # dark royal blue
  "3"  = "#1A459F",  # royal blue
  "4"  = "#2563C4",  # strong blue
  "5"  = "#3F86DB",  # sky blue 
  "6"  = "#4A9C8A",  # muted teal
  "7"  = "#5BA777",  # muted green
  "8"  = "#78B067",  # soft green
  "9"  = "#9BB95E",  # olive green
  "10" = "#BCC45C"   # muted yellow-green
)

# Define boundary info
bin_info <- school_data |>
  summarise(
    .by = weging_bin,
    lo = min(schoolweging), hi = max(schoolweging), n_schools = n()
  ) |>
  mutate(
    x = as.integer(weging_bin),
    txt = if_else(x <= 6, "white", "black"),
    range = paste(round(lo, 1), round(hi, 1), sep = "-")
  )

# Define boundary colours
bin_line_colour      <- "grey45"
trim_line_colour     <- "red"
outside_point_colour <- "grey75"

# Preserve full dataset
bin_labels <- setNames(
  paste0(
    "Bin ", bin_info$weging_bin,
    " (n=", bin_info$n_schools, ")"
  ),
  as.character(bin_info$weging_bin)
)

# Boundaries halfway between adjacent ntile() bins
bin_boundaries <- school_data |>
  group_by(weging_bin) |>
  summarise(
    bin_min = min(schoolweging),
    bin_max = max(schoolweging),
    .groups = "drop"
  ) |>
  arrange(weging_bin) |>
  mutate(cutoff = (bin_max + lead(bin_min)) / 2) |>
  filter(!is.na(cutoff))

# Plotting
ggplot(
  school_data_full,
  aes(x = schoolweging, y = spreiding)
) +
  geom_point(
    colour = outside_point_colour,
    alpha = 0.5,
    size = 0.5
  ) +
  geom_point(
    data = school_data,
    aes(colour = weging_bin),
    alpha = 0.5,
    size = 0.5
  ) +
  geom_vline(
    data = bin_boundaries,
    aes(xintercept = cutoff),
    colour = bin_line_colour,
    linetype = "dotted",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = limits,
    colour = trim_line_colour,
    linetype = "dashed",
    linewidth = 0.6
  ) +
  scale_colour_manual(
    values = bin_colours,
    labels= bin_labels,
    name = "Schoolweging bin",
    drop = FALSE
  ) +
  guides(
    colour = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      ncol = 5,              
      byrow = TRUE,
      override.aes = list(
        size = 2,            
        alpha = 1
      )
    )
  ) +
  labs(
    title = paste(
      "Average socio-economic disadvantage vs. within-school",
      "socio-economic heterogeneity: explanation of the trimming and grouping"
    ),
    x = "Avg. socio-economic disadvantage",
    y = "Socio-economic heterogeneity"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.title = element_text(size = 10),
    strip.text = element_text(face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 8, face = "italic"),
    legend.text = element_text(size = 7),
    legend.key.width = grid::unit(0.2, "cm"),
    legend.key.height = grid::unit(0.2, "cm")
  )
```

<u>Description</u>

We look at a simple scatter plot of average socio-economic disadvantage and socio-economic heterogeneity within a school. Cutoff points were set at 10% on each side of the graph (red lines), and schools outside these cutoffs were excluded. This was done because we wanted to compare schools with roughly the same average socio-economic disadvantage, but with different within-school heterogeneity; outside the cutoffs, there were too few comparable schools spanning both low and high levels of heterogeneity. Furthermore, this range was split into 10 bins containing approximately the same number of schools (~ n = 445). We will use these bins to see how heterogeneity is associated with HAVO+ advice across different levels of average disadvantage (= bins).

## **Statistical Modelling**

```{r regression_models, include=FALSE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# Regression models

# Model 1
model_1 <- lm(
  HAVO_higher ~ spreiding_c * weging_bin, data = school_data)
summary(model_1)

# Model 2
model_2 <- lm(HAVO_higher ~ spreiding_c * weging_trim_c, data = school_data)
summary(model_2)
```

The output of the statistical models can be seen in the rmd-file.

<u>Description</u>

Model 1 is a linear regression with the percentage of HAVO+ advice as the outcome and centered within-school heterogeneity, average disadvantage grouped into ten bins (a factor), and their interaction as predictors. We used these bin-specific slopes and fitted values for the visualizations; the adjusted R² was 0.181, meaning it explained approximately 18% of the variation in HAVO+ advice. Model 2, which used continuous rather than grouped average disadvantage, produced similar results, suggesting that grouping did not substantially reduce the variance explained.

<u>Model diagnostics</u>

In a subsequent step we checked the diagnostics of our model using residual plots:

```{r regression_assumptions, fig.width=11, fig.height=7, echo=FALSE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------------------------
# Regression assumptions

# Check regression assumptions for both models with appropriate plots
old_par <- par(mfrow = c(2, 2))
plot(model_1, which = c(1, 2, 3, 5), ask = FALSE)
par(old_par)
# old_par <- par(mfrow = c(2, 2))
# plot(model_2, which = c(1, 2, 3, 5), ask = FALSE)
# par(old_par)
```

The plots show no clear pattern in the residuals, a fairly stable spread, and no observations that appear strongly influential, although a few schools have higher leverage. The Q–Q plot shows some (likely non-severe) deviations from normality at the tails.

## **Final visualization and conclusion**

Here our final visualization for the research question is shown: 

```{r final-plot, fig.width=14, fig.height=9.8, dpi=200, out.width="100%", fig.align="center", echo=FALSE, warning=FALSE, message=FALSE}
# ------------------------------------------------------------------------------
# Final visualization - shared settings

# Labels for the axis
het <- "Socio-economic heterogeneity"
adv <- "HAVO+ advice"
dis <- "Avg. socio-economic disadvantage"

# Shared legend style, so that all legends have the same text and key size
legend_theme <- theme(
  legend.position = "bottom",
  legend.title = element_text(size = 7.5),
  legend.text = element_text(size = 7.5),
  legend.key.height = unit(0.5, "cm"),
  legend.box.spacing = unit(4, "pt")
)

# ------------------------------------------------------------------------------
# Final visualization plot - bottom left

# Bin-specific slopes (of the centered heterogeneity, model 1) with 95% CIs
slope_data <- marginaleffects::avg_slopes(
  model_1, variables = "spreiding_c", by = "weging_bin"
)

# Fitted lines with 95% bands over each bin's observed range
spreiding_mean <- mean(school_data$spreiding)
pred_data <- school_data |>
  reframe(
    .by = weging_bin,
    spreiding = seq(min(spreiding), max(spreiding), length.out = 50)
  ) |>
  mutate(spreiding_c = spreiding - spreiding_mean) |>
  marginaleffects::predictions(model = model_1, newdata = _)

# Mean advice per bin: a flat line at this level means "no relationship"
bin_means <- school_data |>
  summarise(.by = weging_bin, mean_advice = mean(HAVO_higher))

# Legend text for the dashed lines (used in both bottom plots)
zero_label <- "No relationship (zero slope)"

# Create the bottom left plot (x-axis: centered heterogeneity from model 1):
p_left <- ggplot(school_data, aes(x = spreiding_c, y = HAVO_higher)) +
  geom_segment(
    data = bin_means,
    aes(
      x = -2.6, xend = 2.6,
      y = mean_advice, yend = mean_advice,
      linetype = zero_label
    ),
    colour = "grey30", linewidth = 0.5
  ) +
  geom_ribbon(
    data = pred_data,
    aes(
      y = estimate, ymin = conf.low, ymax = conf.high,
      fill = weging_bin
    ),
    alpha = 0.25
  ) +
  geom_line(
    data = pred_data,
    aes(y = estimate, colour = weging_bin),
    linewidth = 0.9
  ) +
  facet_wrap(
    ~ weging_bin, ncol = 5,
    labeller = labeller(weging_bin = ~ paste("Bin", .x))
  ) +
  scale_colour_manual(values = bin_colours, guide = "none") +
  scale_fill_manual(values = bin_colours, guide = "none") +
  scale_linetype_manual(name = NULL, values = "dashed") +
  guides(
    linetype = guide_legend(
      theme = theme(legend.key.width = unit(1.2, "cm"))
    )
  ) +
  scale_y_continuous(
    breaks = seq(0, 100, 25),
    labels = scales::label_percent(scale = 1)
  ) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    title = "Fitted HAVO+ advice by socio-economic heterogeneity",
    x = paste(het, "(centered)"), y = adv
  ) +
  theme_minimal() +
  legend_theme +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.margin = margin(0, 5.5, 5.5, 5.5),
    axis.title = element_text(size = 10),
    strip.text = element_text(face = "bold")
  )

# ------------------------------------------------------------------------------
# Final visualization plot - bottom right

# Data for the dashed zero line, so that it gets its own legend entry
zero_line <- data.frame(
  yintercept = 0,
  explanation = "If CI includes zero = slope not significant"
)

# Vertical extent of the highlight box: all CIs of bins 1-5 plus a small margin
box_data <- slope_data |>
  filter(as.integer(weging_bin) <= 5) |>
  summarise(ymin = min(conf.low) - 0.25, ymax = max(conf.high) + 0.25)

# Explanatory text for the highlight box
box_text <- paste(
  "Significant slopes within",
  "bins 1-5: higher heterogeneity",
  "is associated with more",
  "students advised to HAVO+",
  sep = "\n"
)

# Create the bottom right plot:
p_right <- ggplot(
  slope_data,
  aes(x = weging_bin, y = estimate, colour = weging_bin)
) +
  geom_hline(
    data = zero_line,
    aes(yintercept = yintercept, linetype = explanation),
    colour = "grey50", inherit.aes = FALSE
  ) +
  geom_rect(
    data = box_data,
    aes(xmin = 0.55, xmax = 5.88, ymin = ymin, ymax = ymax),
    fill = NA, colour = "lightskyblue", linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_pointrange(
    aes(ymin = conf.low, ymax = conf.high),
    linewidth = 0.8, size = 0.6
  ) +
  geom_text(
    aes(label = sprintf("%+.1f pp", estimate)),
    colour = "black", fontface = "bold",
    hjust = 0, nudge_x = 0.1, size = 2.5
  ) +
  annotate(
    "text", x = 6.3, y = 5.2, label = box_text, hjust = 0, vjust = 1,
    fontface = "bold", size = 3.8, lineheight = 0.95, colour = "black"
  ) +
  annotate(
    "curve", x = 7, y = 3.95, xend = 5.98, yend = 3.3,
    curvature = -0.3, linewidth = 0.5, colour = "black",
    arrow = arrow(length = unit(0.2, "cm"), type = "closed")
  ) +
  scale_x_discrete(expand = expansion(add = c(0.6, 1.2))) +
  scale_colour_manual(values = bin_colours, guide = "none") +
  scale_linetype_manual(name = NULL, values = "dashed") +
  guides(
    linetype = guide_legend(
      theme = theme(legend.key.width = unit(1.2, "cm"))
    )
  ) +
  scale_y_continuous(
    breaks = seq(-2, 6, 2),
    labels = scales::label_number(style_positive = "plus", suffix = " pp")
  ) +
  coord_cartesian(ylim = c(-2, 6)) +
  labs(
    title = "Slopes with confidence intervals of heterogeneity on advice",
    x = paste(dis, "(Bin)"),
    y = paste0("Change in ", adv, " per unit of ", tolower(het))
  ) +
  theme_minimal() +
  legend_theme +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.margin = margin(0, 5.5, 5.5, 5.5),
    axis.title = element_text(size = 10)
  )

# ------------------------------------------------------------------------------
# Final visualization - plot title and subtitle at the top:

# Define title and subtitle:
final_title <- paste(
  "More socio-economically mixed schools give more HAVO+ advice,",
  "especially when disadvantage is lower"
)
final_subtitle <- paste(
  "How is within-school heterogeneity in socio-economic disadvantage",
  "associated with the proportion of students receiving final",
  "HAVO-or-higher advice, accounting for the school population’s",
  "average socio-economic disadvantage?"
)

# Footnote at the bottom: model description and data sources
model_note <- paste0(
  "Model: Unweighted linear regression model with interaction term; ",
  "slopes are estimated separately within each avg. disadvantage bin; ",
  "socio-economic heterogeneity was centered; ",
  "95% confidence intervals, adj. R\u00b2 = .181 (p < .001)"
)
source_note <- paste0(
  "Data sources: DUO (Dienst Uitvoering Onderwijs) open education data, ",
  "schooladviezen 2024\u20132025 (duo.nl/open_onderwijsdata); ",
  "Onderwijsinspectie (Dutch Education Inspectorate), schoolweging ",
  "2024\u20132025, calculated by Statistics Netherlands" ,
  "(onderwijsinspectie.nl)"
)
final_caption <- paste(model_note, source_note, sep = "\n")

# ------------------------------------------------------------------------------
# Final visualization legend bins

# Height of the legend panel in cm 
top_cm <- 2.7
# Distance in cm between the legend box and the plots below
gap_cm <- 0.1
# Approximate cm per y-unit of the panel (y runs from -1 to 1)
cm_per_y <- (top_cm - 0.15) / 2

# Strip geometry (x: 0-100 across the panel; y: -1 to 1)
strip_x0 <- 44.5
tile_w <- 5.2
tile_h <- 0.76
tile_yc <- 0.13
y_head <- 0.66
y_top <- 0.78
y_bottom <- -0.78

# "Bin X" size equals the facet strip titles in the left plot
bin_font <- 8.8 / ggplot2::.pt

# Bin summary: range of average disadvantage and number of schools per bin
bin_info <- school_data |>
  summarise(
    .by = weging_bin,
    lo = min(schoolweging), hi = max(schoolweging), n_schools = n()
  ) |>
  mutate(
    x = as.integer(weging_bin),
    tile_x = strip_x0 + (x - 0.5) * tile_w,
    range = paste(round(lo, 1), round(hi, 1), sep = "-")
  )

# Two note rows below the strip: bin ranges, schools per bin, trimming
strip_note <- paste0(
  "Numbers within bins show the range of avg. socio-economic disadvantage. ",
  paste(unique(range(bin_info$n_schools)), collapse = "-"),
  " schools per bin.\n",
  "Schools in the lowest and highest disadvantage deciles excluded ",
  "due to lower within-school socio-economic heterogeneity."
)

# Explanation box: bold term (left column) + short sentence (right column)
box_font <- 3
box_items <- data.frame(
  term = c(
    "HAVO+ advice:",
    "Avg. socio-economic\ndisadvantage:",
    "Socio-economic\nheterogeneity:"
  ),
  text = c(
    "Share of final recommendations that are HAVO, HAVO/VWO or VWO.",
    paste(
      "Average socio-economic disadvantage of the school population;",
      "higher = more disadvantaged. Defines the bins."
    ),
    paste(
      "Differences in socio-economic disadvantage among pupils",
      "within a school; higher = more heterogeneous."
    )
  )
)
box_items$text <- stringr::str_wrap(box_items$text, width = 60)

# Spread the items evenly over the same height as the right-hand strip
line_h <- box_font * ggplot2::.pt * 1.2 * 0.95 / 72 * 2.54 / cm_per_y
box_items$n_lines <- pmax(
  stringr::str_count(box_items$text, "\n"),
  stringr::str_count(box_items$term, "\n")
) + 1
text_h <- sum(box_items$n_lines) * line_h
item_gap <- max((y_top - y_bottom - text_h) / (nrow(box_items) - 1), 0.04)
box_items$y <- y_top -
  c(0, cumsum(head(box_items$n_lines * line_h + item_gap, -1)))

# Create the legend panel: one box around explanation and bin strip
p_legend <- ggplot(bin_info) +
  annotate(
    "rect",
    xmin = 1.5, xmax = 98.5, ymin = -0.94, ymax = 0.94,
    fill = "grey97", colour = "grey30", linewidth = 0.3
  ) +
  # explanation text
  geom_text(
    data = box_items,
    aes(x = 3.5, y = y, label = term),
    hjust = 0, vjust = 1, fontface = "bold", size = box_font,
    lineheight = 0.95, colour = "grey15"
  ) +
  geom_text(
    data = box_items,
    aes(x = 14.5, y = y, label = text),
    hjust = 0, vjust = 1, size = box_font, lineheight = 0.95,
    colour = "grey15"
  ) +
  # compressed bin strip (white text in all bins)
  geom_tile(
    aes(x = tile_x, y = tile_yc, fill = weging_bin),
    width = tile_w * 0.94, height = tile_h
  ) +
  geom_text(
    aes(x = tile_x, y = tile_yc + 0.15, label = paste("Bin", x)),
    colour = "white", fontface = "bold", size = bin_font
  ) +
  geom_text(
    aes(x = tile_x, y = tile_yc - 0.15, label = range),
    colour = "white", size = 2.6
  ) +
  annotate(
    "text", x = strip_x0 + 5 * tile_w, y = y_head, fontface = "bold",
    size = 3.3, colour = "grey15",
    label = "10 equal-sized bins of avg. socio-economic disadvantage"
  ) +
  annotate(
    "text", x = strip_x0, y = y_head, hjust = 0,
    size = 3.1, colour = "grey15", label = "\u2190"
  ) +
  annotate(
    "text", x = strip_x0 + 1.2, y = y_head, hjust = 0,
    fontface = "italic", size = box_font, colour = "grey15",
    label = "Less disadvantaged"
  ) +
  annotate(
    "text", x = strip_x0 + 10 * tile_w, y = y_head, hjust = 1,
    size = 3.1, colour = "grey15", label = "\u2192"
  ) +
  annotate(
    "text", x = strip_x0 + 10 * tile_w - 1.2, y = y_head, hjust = 1,
    fontface = "italic", size = box_font, colour = "grey15",
    label = "More disadvantaged"
  ) +
  annotate(
    "text", x = strip_x0 + 5 * tile_w, y = -0.58, size = 2.7,
    lineheight = 0.95, colour = "grey30", label = strip_note
  ) +
  scale_fill_manual(values = bin_colours) +
  coord_cartesian(xlim = c(0, 100), ylim = c(-1, 1), expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(2, 0, 2, 0)
  )

# ------------------------------------------------------------------------------
# Final plot

# Bottom row: scatterplot (left), slope plot (right)
bottom_row <- wrap_elements(full = p_left) | wrap_elements(full = p_right)

# Create final plot
final_plot <- wrap_elements(full = p_legend) / plot_spacer() / bottom_row +
  plot_layout(heights = unit(c(top_cm, gap_cm, 1), c("cm", "cm", "null"))) +
  plot_annotation(
    title = final_title,
    subtitle = final_subtitle,
    caption = final_caption,
    theme = theme(
      plot.title = element_text(face = "bold", size = 18),
      plot.subtitle = element_text(
        colour = "grey30", size = 9.2, margin = margin(b = 13)
      ),
      plot.caption = element_text(
        colour = "grey40", size = 7.5, hjust = 0, lineheight = 1.1,
        margin = margin(t = 1.5)   
      )
    )
  )
final_plot
```

<u>Short description of the final visualization</u>

The final visualization groups schools into ten bins of average socio-economic disadvantage. The left panel shows how fitted HAVO+ advice shares vary with within-school heterogeneity, while the right panel shows the regression slope for each bin; both include 95% confidence intervals. Showing these plots together helps readers compare the relationships across bins and see both their size and uncertainty. With the explanation box, the notes at the bottom of the graphic, and the titles and subtitles, we have tried to include all the necessary information so that the graphic can be understood (for example, as a research poster at a research conference) without the need for further explanations or additional information. 

<u>Conclusion</u>

Within the middle range of average socio-economic disadvantage that we examine, greater within-school heterogeneity is associated with a higher proportion of students receiving final HAVO-or-higher advice, mainly among relatively less disadvantaged schools (bins 1–5). In these bins, a one-unit increase in heterogeneity is positively associated with approximately 2.5–3.5-percentage-point higher share of HAVO+ advice. The association is non-significant or generally weaker in more disadvantaged schools, although it remains positive in bin 10. These findings suggest that the relationship between socio-economic heterogeneity and educational advice varies with schools average disadvantage. 

<u>Limitations on interpretation</u>

It is important to mention that this is not evidence for a causal effect. While our analysis identifies a relationship, the estimated effect sizes and the differences between bins are not, on their own, sufficient to support policy recommendations. The findings may also depend on the replacement of suppressed counts and the trimming and binning choices. Furthermore, school-level heterogeneity may also obscure classroom-level effects, where mechanisms such as peer interactions and teaching practices are likely to have a more direct effect. These mechanisms are complex and may depend on how students are grouped and supported; a heterogeneous school population does not necessarily imply heterogeneous classrooms. Nevertheless, our findings are compatible with the hypothesis that socio-economic diversity may benefit the further educational opportunities of both advantaged and disadvantaged students.

## **AI use**

While working on this project, we utilized artificial intelligence at various stages, particularly when creating the code for the final plot, but also for creating exploratory plots and other small code snippets and general feedback. Throughout the process, we always provided as clear instructions as possible regarding exactly how the visualization should look and how the code should be written, and we had the system generate only small sections step by step, one after another, so that we could review and verify each code section in R. Through iterative feedback processes, we repeatedly had the visualization adjusted or made manual changes ourselves. We realized how important our own understanding of the data and the codebook was for the quality of the output, especially for the correct data preparation. We also realized that it is crucial for us to provide very clear guidance and describe in great detail exactly what we want. As input, we provided the AI with the codebook, the assignment objectives, and, depending on the phase, our code to date. Overall, AI has made our work much easier, and the quality of the results was very good (depending on the precision of our prompts and the clarity of our objectives). AI was also very helpful as a feedback partner for ideas and brainstorming.

Filip worked primarily with the Copilot agent in the terminal, and Henning worked primarily with the latest Claude model (claude-sonnet-5.5) in the UvA AI Chat. 

