---
title: "DV Assignment 2, Part 1: [Your Title Here]"
author: "Group X: Firstname Lastname, Firstname Lastname, Firstname Lastname"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

<!--
HOW TO KNIT THIS FILE

1. Open your group project in RStudio first (the `Project:` button in the
   top right should show your repo name). Keep this file in `report/`.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd in `report/`.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below sources scripts/00-packages.R and
   scripts/01-get-data.R, so packages and data are always loaded. The first
   knit downloads the data, so you need an internet connection.

If knitting fails:
- "could not find function": the package isn't loaded - add it to
  scripts/00-packages.R.
- "object not found": you created it in the Console, not in this file (or
  in a script that this file doesn't source).
- "cannot open file": check that the project is open, and use
  here("folder", "file") for paths instead of typing them out.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))
```

## Research question

*What is your research question? Is it a question of analytics, inference or prediction? Who is the target audience for your final visualization, and how does that shape your choices?*

```{r setup, include = FALSE}
#Set up of the variables




# The ODS identifies locations by institution code plus a different location
# code. Average locations to institution level using pupil counts as weights.
weging_by_school <- schoolweging |>
  mutate(INSTELLINGSCODE = str_extract(OVT, "^[^|]+")) |>
  filter(
    !is.na(INSTELLINGSCODE),
    !is.na(schoolweging),
    !is.na(spreiding),
    !is.na(aantal_leerlingen),
    aantal_leerlingen > 0
  ) |>
  group_by(INSTELLINGSCODE) |>
  summarise(
    schoolweging = weighted.mean(schoolweging, aantal_leerlingen),
    spreiding = weighted.mean(spreiding, aantal_leerlingen),
    .groups = "drop"
  )

# STANDARDAZING TEST SCORES
# Provider is used only to standardize otherwise incomparable score scales.
scores_by_provider <- eindscores |>
  filter(SOORT_PO == "Bo") |>
  pivot_longer(
    cols = matches("^(IEP|ROUTE8|DIA|AMN|DOE|LIB)_(AANTAL|GEM)$"),
    names_to = c("provider", ".value"),
    names_pattern = "^(.+)_(AANTAL|GEM)$"
  ) |>
  mutate(
    AANTAL = as.numeric(
      if_else(
        str_detect(AANTAL, "^[0-9]+$"),
        AANTAL,
        NA_character_
      )
    )
  ) |>
  filter(AANTAL > 0, !is.na(GEM)) |>
  group_by(INSTELLINGSCODE, provider) |>
  summarise(
    mean_score = weighted.mean(GEM, AANTAL),
    tested_pupils = sum(AANTAL),
    .groups = "drop"
  ) |>
  group_by(provider) |>
  mutate(
    score_z = (mean_score - mean(mean_score)) / sd(mean_score)
  ) |>
  ungroup() |>
  filter(is.finite(score_z))

# Combine any multiple provider results into one school-level value, weighted
# by tested pupils. Provider names are not used in the visualizations.
scores_by_school <- scores_by_provider |>
  group_by(INSTELLINGSCODE) |>
  summarise(
    standardized_score = weighted.mean(score_z, tested_pupils),
    .groups = "drop"
  )
plot_data <- scores_by_school |>
  inner_join(weging_by_school, by = "INSTELLINGSCODE") |>
  filter(
    is.finite(standardized_score),
    is.finite(schoolweging),
    is.finite(spreiding)
  )

tertile_cutpoints <- quantile(
  plot_data$schoolweging,
  probs = c(1 / 3, 2 / 3),
  na.rm = TRUE
)
plot_data <- plot_data |>
  mutate(
    disadvantage_level = case_when(
      schoolweging <= tertile_cutpoints[[1]] ~ "Lower disadvantage",
      schoolweging <= tertile_cutpoints[[2]] ~ "Middle disadvantage",
      TRUE ~ "Higher disadvantage"
    ),
    disadvantage_level = factor(
      disadvantage_level,
      levels = c(
        "Lower disadvantage",
        "Middle disadvantage",
        "Higher disadvantage"
      )
    )
  )
```

##Exploratory process
```{r }

plot_score_by_spread <- ggplot(
  plot_data,
  aes(x = spreiding, y = standardized_score)
) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
  facet_wrap(vars(disadvantage_level)) +
  labs(
    title = "Relative scores by spreiding across disadvantage levels",
    subtitle = paste(
      "Schoolweging groups are data-derived tertiles;",
      "each point is one school"
    ),
    x = "Spreiding (variation in disadvantage)",
    y = "Standardized test score (provider-specific z-score)"
  )
print(plot_score_by_spread)
```



```{r setup, include = FALSE}
# ── 1. Schoolweging ───────────────────────────────────────────────────────────
# Extract INSTELLINGSCODE from OVT (format "00AP|C1") and aggregate to school
# level, weighting by number of pupils.
weging <- schoolweging |>
  mutate(INSTELLINGSCODE = str_extract(OVT, "^[^|]+")) |>
  filter(!is.na(schoolweging), !is.na(spreiding), aantal_leerlingen > 0) |>
  group_by(INSTELLINGSCODE) |>
  summarise(
    schoolweging = weighted.mean(schoolweging, aantal_leerlingen),
    spreiding    = weighted.mean(spreiding,    aantal_leerlingen),
    .groups = "drop"
  )

# ── 2. School advice → pct_havo_plus ─────────────────────────────────────────
# Replace "<5" with 2.5, convert to numeric, then compute the share of
# students advised for HAVO or higher.
parse_count <- function(x) {
  as.numeric(if_else(x == "<5", "2.5", x))
}

adviezen <- schooladviezen |>
  filter(SOORT_PO == "Bo") |>
  mutate(across(VSO:ADVIES_NIET_MOGELIJK, parse_count)) |>
  mutate(
    total     = rowSums(across(VSO:ADVIES_NIET_MOGELIJK), na.rm = TRUE),
    havo_plus = HAVO + HAVO_VWO + VWO
  ) |>
  filter(total > 0) |>
  mutate(pct_havo_plus = havo_plus / total) |>
  select(INSTELLINGSCODE, pct_havo_plus)

# ── 3. Join ───────────────────────────────────────────────────────────────────
plot_data2 <- adviezen |>
  inner_join(weging, by = "INSTELLINGSCODE") |>
  filter(
    is.finite(pct_havo_plus),
    is.finite(schoolweging),
    is.finite(spreiding)
  ) |>
  mutate(
    spreiding_q = cut(
      spreiding,
      breaks = quantile(spreiding, probs = 0:4 / 4, na.rm = TRUE),
      labels = c("Q1 (low spreiding)", "Q2", "Q3", "Q4 (high spreiding)"),
      include.lowest = TRUE
    )
  )

```

## Exploratory process

*Describe how you explored the data. Include at least one exploratory visualization below, and a short note on what it shows, and what you learned from it about your research question.*

```{r}
p3 <- ggplot(plot_data2, aes(x = schoolweging, y = pct_havo_plus)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  facet_wrap(vars(spreiding_q)) +
  labs(
    title = "Option 3: Faceted scatter by spreiding quartile",
    x = "Schoolweging", y = "Share advised HAVO or higher"
  )
print(p3)
```

---

**Before submitting:** run the lintr check on this file, knit it, and push your work to GitHub.
