---
title: "DV Assignment 2, Part 1: Effects of school disadvantage heterogeneity"
author: "Group 3: Filip Hlavinka, Henning Lang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

<!--
HOW TO KNIT THIS FILE

1. Open your group project in RStudio first (the `Project:` button in the
   top right should show your repo name). Keep this file in `report/`.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd in `report/`.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below sources scripts/00-packages.R and
   scripts/01-get-data.R, so packages and data are always loaded. The first
   knit downloads the data, so you need an internet connection.

If knitting fails:
- "could not find function": the package isn't loaded - add it to
  scripts/00-packages.R.
- "object not found": you created it in the Console, not in this file (or
  in a script that this file doesn't source).
- "cannot open file": check that the project is open, and use
  here("folder", "file") for paths instead of typing them out.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here::here("scripts", "01-get-data.R"))
```

## **Research question and audience**

<u>Introduction</u>

We have decided to look at how spreiding (within-school variance of disadvantage) could affect test scores/final school advise across different levels of schoolweging (mean school disadvantage). We thought this would be interesting to see if more heterogeneous groups with regard to advantaged/disadvantaged backgrounds (based on parental income and education) could be related to better performance. There is some evidence to suggest that the performance of students can benefit in more heterogeneous social environments, so we thought it would be interesting to analyze if we could observe a similar phenomenon in this data. We made several exploratory plots included in our test scripts folders but decided to keep only a couple of those here for simplicity. At this point, the question is analytical, so we just look at the visualizations to see if there are any discernible patterns which we could later model. The audience could be policymakers and school boards, but given the severe limitations of what we are measuring, a certain level of nuance and expertise is required when interpreting the data.

<u>Short summary</u>

|   |   |
|---|--------|
| `Research question`   | *RQ5 (Own RQ)*: Either **Does within-school heterogeneity in students' socioeconomic disadvantage predict students' test scores after controlling for the school's average level of student socioeconomic disadvantage?** or **Does within-school heterogeneity in students' socioeconomic disadvantage predict the share of students receiving a HAVO-or-higher secondary-school advice after controlling for the school's average level of student socioeconomic disadvantage?** | 
| `Analytical approach` | Analytical | 
| `Target audience`     | Policymakers and school boards interested in how school composition affects educational opportunity (But with caution and some limitations when interpreting it). |


```{r variables_data, include = FALSE}
# Set up of the variables

# code. Average locations to institution level using pupil counts as weights.
weging_by_school <- schoolweging |>
  mutate(INSTELLINGSCODE = str_extract(OVT, "^[^|]+")) |>
  filter(
    !is.na(INSTELLINGSCODE),
    !is.na(schoolweging),
    !is.na(spreiding),
    !is.na(aantal_leerlingen),
    aantal_leerlingen > 0
  ) |>
  group_by(INSTELLINGSCODE) |>
  summarise(
    schoolweging = weighted.mean(schoolweging, aantal_leerlingen),
    spreiding = weighted.mean(spreiding, aantal_leerlingen),
    .groups = "drop"
  )

# STANDARDAZING TEST SCORES
# Provider is used only to standardize otherwise incomparable score scales.
scores_by_provider <- eindscores |>
  filter(SOORT_PO == "Bo") |>
  pivot_longer(
    cols = matches("^(IEP|ROUTE8|DIA|AMN|DOE|LIB)_(AANTAL|GEM)$"),
    names_to = c("provider", ".value"),
    names_pattern = "^(.+)_(AANTAL|GEM)$"
  ) |>
  mutate(
    AANTAL = as.numeric(
      if_else(
        str_detect(AANTAL, "^[0-9]+$"),
        AANTAL,
        NA_character_
      )
    )
  ) |>
  filter(AANTAL > 0, !is.na(GEM)) |>
  group_by(INSTELLINGSCODE, provider) |>
  summarise(
    mean_score = weighted.mean(GEM, AANTAL),
    tested_pupils = sum(AANTAL),
    .groups = "drop"
  ) |>
  group_by(provider) |>
  mutate(
    score_z = (mean_score - mean(mean_score)) / sd(mean_score)
  ) |>
  ungroup() |>
  filter(is.finite(score_z))

# Combine any multiple provider results into one school-level value, weighted
# by tested pupils. Provider names are not used in the visualizations.
scores_by_school <- scores_by_provider |>
  group_by(INSTELLINGSCODE) |>
  summarise(
    standardized_score = weighted.mean(score_z, tested_pupils),
    .groups = "drop"
  )
plot_data <- scores_by_school |>
  inner_join(weging_by_school, by = "INSTELLINGSCODE") |>
  filter(
    is.finite(standardized_score),
    is.finite(schoolweging),
    is.finite(spreiding)
  )

tertile_cutpoints <- quantile(
  plot_data$schoolweging,
  probs = c(1 / 3, 2 / 3),
  na.rm = TRUE
)
plot_data <- plot_data |>
  mutate(
    disadvantage_level = case_when(
      schoolweging <= tertile_cutpoints[[1]] ~ "Lower disadvantage",
      schoolweging <= tertile_cutpoints[[2]] ~ "Middle disadvantage",
      TRUE ~ "Higher disadvantage"
    ),
    disadvantage_level = factor(
      disadvantage_level,
      levels = c(
        "Lower disadvantage",
        "Middle disadvantage",
        "Higher disadvantage"
      )
    )
  )
```


## **Exploratory plots**


**Plot 1**

```{r ex_plot1, echo=FALSE, message=FALSE, warning=FALSE}
plot_data3 <- scores_by_provider |>
  inner_join(weging_by_school, by = "INSTELLINGSCODE") |>
  filter(
    is.finite(mean_score),
    is.finite(schoolweging),
    is.finite(spreiding)
  )
plot_spread_and_weight <- ggplot(
  plot_data,
  aes(x = schoolweging, y = spreiding)
) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(
    title = "Schoolweging and spreiding",
    subtitle = "Each point represents one matched regular primary school",
    x = "Schoolweging (overall disadvantage)",
    y = "Spreiding (variation in disadvantage)"
  )
print(plot_spread_and_weight)
```

<u>Description</u>

The above graph visualizes the variance of disadvantage within school (y-axis) and different levels of overall disadvantage of the school (x-axis). Each point represents an individual institution. After seeing this, we thought it would be interesting to then see if schools in roughly the same schoolweging would have different performance scores depending on their spreiding.


**Plot 2**

```{r ex_plot2, echo=FALSE, message=FALSE, warning=FALSE}
plot_score_by_spread <- ggplot(
  plot_data,
  aes(x = spreiding, y = standardized_score)
) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
  facet_wrap(vars(disadvantage_level)) +
  labs(
    title = "Relative scores by spreiding across disadvantage levels",
    subtitle = paste(
      "Schoolweging groups are data-derived tertiles;",
      "each point is one school"
    ),
    x = "Spreiding (variation in disadvantage)",
    y = "Standardized test score (provider-specific z-score)"
  )
print(plot_score_by_spread)
```

<u>Description</u>

The above graph uses standardized test scores (as schools use different providers) to see if student performance is higher with high or low variance of disadvantage. This is done for three levels of overall disadvantage: Low, Mid, and High. There doesn't seem to be any apparent trend in the relationship we expected. However, we can also see that, overall, students with higher disadvantage also have overall worse performance.


```{r setup2, include = FALSE}
# ── 1. Schoolweging ───────────────────────────────────────────────────────────
# Extract INSTELLINGSCODE from OVT (format "00AP|C1") and aggregate to school
# level, weighting by number of pupils.
weging <- schoolweging |>
  mutate(INSTELLINGSCODE = str_extract(OVT, "^[^|]+")) |>
  filter(!is.na(schoolweging), !is.na(spreiding), aantal_leerlingen > 0) |>
  group_by(INSTELLINGSCODE) |>
  summarise(
    schoolweging = weighted.mean(schoolweging, aantal_leerlingen),
    spreiding = weighted.mean(spreiding, aantal_leerlingen),
    .groups = "drop"
  )

# ── 2. School advice → pct_havo_plus ─────────────────────────────────────────
# Replace "<5" with 2.5, convert to numeric, then compute the share of
# students advised for HAVO or higher.
parse_count <- function(x) {
  as.numeric(if_else(x == "<5", "2.5", x))
}

adviezen <- schooladviezen |>
  filter(SOORT_PO == "Bo") |>
  mutate(across(VSO:ADVIES_NIET_MOGELIJK, parse_count)) |>
  mutate(
    total     = rowSums(across(VSO:ADVIES_NIET_MOGELIJK), na.rm = TRUE),
    havo_plus = HAVO + HAVO_VWO + VWO
  ) |>
  filter(total > 0) |>
  mutate(pct_havo_plus = havo_plus / total) |>
  select(INSTELLINGSCODE, pct_havo_plus)

# ── 3. Join ───────────────────────────────────────────────────────────────────
plot_data <- adviezen |>
  inner_join(weging, by = "INSTELLINGSCODE") |>
  filter(
    is.finite(pct_havo_plus),
    is.finite(schoolweging),
    is.finite(spreiding)
  ) |>
  mutate(
    spreiding_q = cut(
      spreiding,
      breaks = quantile(spreiding, probs = 0:4 / 4, na.rm = TRUE),
      labels = c("Q1 (low spreiding)", "Q2", "Q3", "Q4 (high spreiding)"),
      include.lowest = TRUE
    )
  )
```


**Plot 3**

```{r plot3, echo=FALSE, message=FALSE, warning=FALSE}
slope_data <- plot_data |>
  mutate(
    weging_q = cut(
      schoolweging,
      breaks = quantile(schoolweging, probs = 0:4 / 4, na.rm = TRUE),
      labels = c("Q1 (least disadvantaged)", "Q2", "Q3",
                 "Q4 (most disadvantaged)"),
      include.lowest = TRUE
    )
  ) |>
  group_by(spreiding_q, weging_q) |>
  summarise(mean_havo = mean(pct_havo_plus), .groups = "drop")
p3 <- ggplot(
  slope_data,
  aes(x = weging_q, y = mean_havo, color = spreiding_q, group = spreiding_q)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_viridis_d(name = "Spreiding quartile") +
  labs(
    title = "School advise by schools' disadvantage mean and heterogeneity",
    x = "Schoolweging quartile", y = "Share advised HAVO or higher"
  )
print(p3)
```

<u>Description</u>

Additionally, we performed the same analysis, but instead of standardized scores, we used the share of students advised to HAVO or higher as outcome (i. e. the final school advise). We used schoolweging quartiles on the x-axis and the spreiding-quartiles as lines in different colours. The graph shows that given the particular effect of schoolweging there is no relevant additional effect of within-school disadvantage heterogeneity. 


**Plot 4**

```{r plot4, echo=FALSE, message=FALSE, warning=FALSE}
p4 <- ggplot(plot_data, aes(x = schoolweging, y = pct_havo_plus)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  facet_wrap(vars(spreiding_q)) +
  labs(
    title = "School advise by schools' disadvantage mean and heterogeneity",
    x = "Schoolweging", y = "Share advised HAVO or higher"
  )
print(p4)
```

<u>Description</u>

In this plot again the share of students advised to HAVO or higher was the considered outcome. The plots are also this time split according to within-school disadvantage variance rather than overall variance, as it was the case above. These visualizations seem to follow the same pattern: there seems to be no aditional effect of spreiding, but definitely an effect of schoolweging.


## **Feedback**

We'd appreciate your **feedback on the second plot**. Thanks!


## **Link to our GitHub repository**

https://github.com/HenningLa7/bds_toolbox_data_viz_group3
